Where Do Events Come From?

### 2018-08-02 paper.tex
...

Interrupts service routines (ISRs) are software entry points that execute in
response to hardware interrupts from peripherals such as timers and GPIOs.
ISRs are at the lowest level interface between hardware and software and are
the absolute source of inputs to programs.
%
Typically, an ISR starts to execute as soon as the hardware interrupt occurs
and suspends the normal program flow abruptly.
Such asynchronous behavior reflects the inherent concurrent nature of
peripherals interacting with the external world.

@@

We propose to extend \CEU with asynchronous ISRs.
However, asynchronous execution confronts the synchronous mindset of \CEU since
the assumption that reactions are atomic no longer holds.
Not only this may lead to race conditions at a fine grain, but may also affect
the ordering of events at a coarse grain: the effect of an earlier event may be
perceived after the effect of a later event.
%
Nevertheless, our goal is to preserve the well-behaved interaction between the
\CEU application and the environment as of Figure~\ref{fig.env}, even in the
presence of asynchronous ISRs as illustrated in Figure~\ref{fig.async}:
%
the original application is now represented by the synchronous \CEU and the
environment by the asynchronous \CEU, all inside the same application.
%
This way, we want to push all subtleties of asynchronous execution into device
drivers which emit input events to regular synchronous code in \CEU exactly
as before.

@@

Listing~\ref{lst.gpio.sync}~and~\ref{lst.gpio.async} implements an application
that uses GPIO to link a button (pin 02) to an LED (pin 13) such that the LED
is on whenever the button is pressed and off whenever it is unpressed.

The synchronous side (Listing~\ref{lst.gpio.sync}) declares its interface with
the external world and includes the asynchronous side as a driver (ln~1--3).
It then starts with the LED off (ln~5) and enters in a loop (ln~6--9) that,
whenever the button changes (ln~7), changes the state of the LED with the new
value (ln~8).

The asynchronous side (Listing~\ref{lst.gpio.async}) implements the output and
input events.
%
An \code{output} implementation (ln~4--6) is similar to a parameterized
subroutine: whenever the application invokes \code{emit}, the output body
executes atomically.
\CEU supports inline C between curly braces (ln~3,5) with interpolation to
evaluates \CEU expressions (e.g., \code{@v}).
This allows drivers to take advantage of existing libraries in embedded
toolchains, such as Arduino.
%
In the example, the driver sets pin 13 as output (ln~3) when it is included,
and sets its state whenever it is emitted (ln~5).
%
An \code{input} event implementation requires an ISR registered with the
\code{spawn async/isr} primitive (ln~16--18), which is automatically invoked
whenever the associated interrupt occurs (e.g., \code{INT0\_vect}).
%
An ISR in \CEU will typically have an \code{emit} to an input event inside its
body to awake the synchronous side.
However, although the ISR executes asynchronously as soon as the interrupt
occurs, the \code{emit} respects the semantics of \CEU and does not interrupt
an ongoing reaction on the synchronous side (going into a queue to be
discussed).
%
In the example, the input driver first configures pin 2 to behave as input and
to generate an external interrupt on level transitions (ln~11--15).

The example illustrates the clear separation between the application and the
driver through an event-driven interface that reflects the desired architecture
of Figures~\ref{fig.env}~and~\ref{fig.async}.
%
Now the whole application is written in \CEU:
    the synchronous side remains well behaved with no low-level calls, and
    the asynchronous side deals with all complexity in device drivers.
%
Nevertheless, drivers are typically write-once components developed by embedded
specialists, which are reused in most applications.
%
Note that each architecture requires a mapping between the \code{async/isr} and
the actual interrupt vector table, which we provide with our system.

As we discuss in the next section, supporting ISRs in the same language and
memory space of the application allows programmers to choose between efficiency
and safety during development by providing (and experimenting with) multiple
interfaces with different levels of abstraction.

- TODO adc

\begin{lstlisting}[
  xleftmargin=1em,
  numbers=left,
  basicstyle=\ttfamily\small,
  float=t,
  caption={XXX.},
  label={lst.gpio.async},
  escapechar=\%,
]
// A/D DRIVER (adc.ceu)

@@

do finalize with
  {
    ADCSRA &= B01111111;  // disables A/D converter
    ACSR    = B10000000;  // disables comparator
    DIDR0  |= B00111111;  // disables pins
  }
end

...     // driver implementation

loop do
    var int v;
    do
        #include "adc.ceu"
        emit ADC_REQUEST(A0);
        v = await ADC_DONE;
    end
    ...     // uses v
    await 1h;
end
\end{lstlisting}

>>>>

Interrupts service routines (ISRs) are software entry points that execute in
response to hardware interrupts from peripherals such as timers and GPIOs.
ISRs are at the lowest interface layer between the hardware and software and
are the absolute source of inputs to programs.
%
An ISR starts to execute as soon as the hardware interrupt occurs, suspending
the ongoing program flow abruptly.
Such asynchronous behavior reflects the inherent concurrent nature of
peripherals interacting with the external world.

@@

We propose to extend \CEU with asynchronous ISRs.
However, asynchronous execution confronts the synchronous mindset of \CEU since
the assumption that reactions are atomic no longer holds.
Not only this may lead to race conditions at a fine grain, but may also affect
the ordering of events at a coarse grain: the effect of an earlier event may be
perceived after the effect of a later event.
%
Still, our goal is to preserve the well-behaved interaction between the \CEU
application and the environment even in the presence of asynchronous ISRs.
Figure~\ref{fig.async} adapts the event loop of Figure~\ref{fig.env} to fit all
inside the same \CEU application, now with the original application represented
as the synchronous side and the environment as the asynchronous side.
%
Our idea is to push all subtleties of asynchronous execution into device
drivers which emit input events to regular synchronous code in \CEU exactly
as before.

@@

As a first example, Listing~\ref{lst.gpio.sync}~and~\ref{lst.gpio.async} uses
GPIOs to virtually connect a button (pin 02) to an LED (pin 13) such that the
LED is on whenever the button is pressed and off whenever it is unpressed.

Listing~\ref{lst.gpio.sync}, the synchronous side, first declares its interface
with the external world and includes the asynchronous side as a driver
(ln~1--3).
It then starts with the LED off (ln~5) and enters in a loop (ln~6--9) that,
whenever the button changes (ln~7), changes the state of the LED with the new
value (ln~8).

Listing~\ref{lst.gpio.async}, the asynchronous side, implements the output and
input events.
%
An \code{output} implementation (ln~4--6) is similar to a parameterized
subroutine: whenever the application invokes \code{emit}, the output body
executes atomically.
\CEU supports inline C between curly braces (ln~3,5) and interpolation to
evaluate \CEU expressions (e.g., \code{@v}).
This allows drivers to take advantage of existing libraries in embedded
toolchains, such as Arduino.
%
In the example, when the driver is included, it configures pin 13 as output
(ln~3) and sets its state whenever \code{PIN\_13} is emitted (ln~5).
%
An \code{input} event implementation uses an ISR registered with the
\code{spawn async/isr} primitive (ln~16--18), which is automatically invoked
whenever the associated interrupt occurs (e.g., \code{INT0\_vect}).
%
An ISR in \CEU will typically emit an input event inside its body to awake the
synchronous side (ln~17).
However, although the ISR executes asynchronously when the interrupt occurs,
the \code{emit} goes into a queue (to be discussed) and does not interrupt the
ongoing (suspended) reaction on the synchronous side.
%
In the example, the input driver also configures pin 2 to behave as input and
to generate external interrupts on level transitions (ln~11--15).

The example illustrates the clear separation between the application and the
driver through an event-driven interface that reflects the desired architecture
of Figures~\ref{fig.env}~and~\ref{fig.async}.
%
Now, the whole application is written in \CEU:
    the synchronous side remains well behaved with no low-level calls, and
    the asynchronous side deals with all complexity of device drivers.
%
Nevertheless, drivers are typically write-once components developed by embedded
specialists, which are reused in most applications.
%
Note that each architecture requires a mapping between the \code{async/isr} and
the actual interrupt vector table, which we provide with our system.

As we discuss in the next section, supporting ISRs in the same language and
memory space of the application allows programmers to choose between efficiency
and safety during development by providing multiple interfaces with different
levels of abstraction.

\begin{lstlisting}[
  xleftmargin=1em,
  numbers=left,
  basicstyle=\ttfamily\small,
  float=t,
  caption={XXX.},
  label={lst.adc},
  escapechar=\%,
]
// A/D DRIVER (adc.ceu)

@@

  do finalize with
    {
      ADCSRA &= B01111111;  // disables A/D converter
      ACSR    = B10000000;  // disables comparator
      DIDR0  |= B00111111;  // disables pins
    }
  end
  ...     // driver initialization
  ...     // input / output implementations

// APPLICATION (main.ceu)

  loop do
    var int v;
    do
      #include "adc.ceu" // driver contents above
      emit ADC_REQUEST(A0);
      v = await ADC_DONE;
    end
    ...                  // uses sensor value "v"
    await 1h;
  end
\end{lstlisting}

The example in Listing~\ref{lst.adc} illustrates the use of a lexically-scoped
driver for the A/D converter.
Note that the driver (ln~1--14) and application (ln~16--27) are actually in
separate files.
%
The application is a typical periodic sensor sampling loop, but which
designates an explicit \code{do--end} scope to execute the driver (ln~21) and
use it (ln~22--23).
%
The driver specifies a finalization code (ln~6--12) which executes
unconditionally whenever it goes out of scope (ln~24).
This code completely disables all A/D functionality, saving energy.
%
As discussed for Listing~\ref{lst.ceu}, finalizers also execute due to
\code{par/or} termination (or other abortion mechanisms of \CEU).
%
Note that the driver interface (ln~3--4) is visible only inside the block and
thus cannot be inadvertently used outside it.
